% Legal notices for IBM-provided documentation about IBM products.  Should
% be included as a separate chapter.

This information was developed for products and services offered in the
US\@.  This material might be available from IBM in other languages.
However, you may be required to own a copy of the product or product
version in that language in order to access it.

IBM may not offer the products, services, or features discussed in this
document in other countries.  Consult your local IBM representative for
information on the products and services currently available in your area.
Any reference to an IBM product, program, or service is not intended to
state or imply that only that IBM product, program, or service may be
used.  Any functionally equivalent product, program, or service that does
not infringe any IBM intellectual property right may be used instead.
However, it is the user's responsibility to evaluate and verify the
operation of any non-IBM product, program, or service.

IBM may have patents or pending patent applications covering subject
matter described in this document.  The furnishing of this document does
not grant you any license to these patents.  You can send license
inquiries, in writing, to:
\begin{quote}
  IBM Director of Licensing\\
  IBM Corporation\\
  North Castle Drive, MD-NC119\\
  Armonk, NY 10504-1785\\
  US
\end{quote}

For license inquiries regarding double-byte character set (DBCS)
information, contact the IBM Intellectual Property Department in your
country or send inquiries, in writing, to:
\begin{quote}
  Intellectual Property Licensing\\
  Legal and Intellectual Property Law\\
  IBM Japan Ltd.\\
  19-21, Nihonbashi-Hakozakicho, Chuo-ku\\
  Tokyo 103-8510, Japan
\end{quote}

INTERNATIONAL BUSINESS MACHINES CORPORATION PROVIDES THIS PUBLICATION ``AS
IS'' WITHOUT WARRANTY OF ANY KIND, EITHER EXPRESS OR IMPLIED, INCLUDING,
BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF NON-INFRINGEMENT,
MERCHANTABILITY OR FITNESS FOR A PARTICULAR PURPOSE\@.  Some jurisdictions
do not allow disclaimer of express or implied warranties in certain
transactions, therefore, this statement may not apply to you.

This information could include technical inaccuracies or typographical
errors.  Changes are periodically made to the information herein; these
changes will be incorporated in new editions of the publication.  IBM may
make improvements and/or changes in the product(s) and/or the program(s)
described in this publication at any time without notice.

Any references in this information to non-IBM websites are provided for
convenience only and do not in any manner serve as an endorsement of those
websites.  The materials at those websites are not part of the materials
for this IBM product and use of those websites is at your own risk.

IBM may use or distribute any of the information you provide in any way it
believes appropriate without incurring any obligation to you.

Licensees of this program who wish to have information about it for the
purpose of enabling: (i)~the exchange of information between independently
created programs and other programs (including this one) and (ii)~the
mutual use of the information which has been exchanged, should contact:
\begin{quote}
  IBM Director of Licensing\\
  IBM Corporation\\
  North Castle Drive, MD-NC119\\
  Armonk, NY 10504-1785\\
  US
\end{quote}

Such information may be available, subject to appropriate terms and
conditions, including in some cases, payment of a fee.

The licensed program described in this document and all licensed material
available for it are provided by IBM under terms of the IBM Customer
Agreement, IBM International Program License Agreement or any equivalent
agreement between us.

The performance data discussed herein is presented as derived under
specific operating conditions.  Actual results may vary.

The client examples cited are presented for illustrative purposes only.
Actual performance results may vary depending on specific configurations
and operating conditions.

The performance data and client examples cited are presented for
illustrative purposes only.  Actual performance results may vary depending
on specific configurations and operating conditions.

Information concerning non-IBM products was obtained from the suppliers of
those products, their published announcements or other publicly available
sources.  IBM has not tested those products and cannot confirm the
accuracy of performance, compatibility or any other claims related to
non-IBM products.  Questions on the capabilities of non-IBM products
should be addressed to the suppliers of those products.

Statements regarding IBM's future direction or intent are subject to
change or withdrawal without notice, and represent goals and objectives
only.

All IBM prices shown are IBM's suggested retail prices, are current and
are subject to change without notice.  Dealer prices may vary.

This information is for planning purposes only.  The information herein is
subject to change before the products described become available.

This information contains examples of data and reports used in daily
business operations.  To illustrate them as completely as possible, the
examples include the names of individuals, companies, brands, and
products.  All of these names are fictitious and any similarity to actual
people or business enterprises is entirely coincidental.

\medskip
COPYRIGHT LICENSE:\\\nopagebreak
This information contains sample application programs in source language,
which illustrate programming techniques on various operating platforms.
You may copy, modify, and distribute these sample programs in any form
without payment to IBM, for the purposes of developing, using, marketing
or distributing application programs conforming to the application
programming interface for the operating platform for which the sample
programs are written.  These examples have not been thoroughly tested
under all conditions.  IBM, therefore, cannot guarantee or imply
reliability, serviceability, or function of these programs.  The sample
programs are provided "AS IS", without warranty of any kind.  IBM shall
not be liable for any damages arising out of your use of the sample
programs.

Each copy or any portion of these sample programs or any derivative work
must include a copyright notice as follows:

\begin{quote}
  © (your company name) (year).\\
  Portions of this code are derived from IBM Corp. Sample Programs.\\
  © Copyright IBM Corp. 2001, 2021.
\end{quote}

\section{Trademarks}

IBM, the IBM logo, and \texttt{ibm.com} are trademarks or registered
trademarks of International Business Machines Corp., registered in many
jurisdictions worldwide. Other product and service names might be
trademarks of IBM or other companies. A current list of IBM trademarks is
available on the web at "Copyright and trademark information" at
\url{www.ibm.com/legal/copytrade.shtml}.

The registered trademark Linux\textregistered{} is used pursuant to a
sublicense from the Linux Foundation, the exclusive licensee of Linus
Torvalds, owner of the mark on a world­wide basis.
